---
title: "Cyclistic_Trips"
author: "Ganesh"
date: "08/08/2021"
output:
  html_document: default
---

#Cyclistic_Trips_Full_Year_Analysis

**Analyzing Cyclistic’s historical trip data to analyze and identify trends**


**Data Source:** Cyclistic is a fictional company. For the purposes of this case study,
the datasets are appropriate and will enable you to answer the business questions. The data has been made available by Motivate International Inc. You can download the data [here](https://divvy-tripdata.s3.amazonaws.com/index.html).

**Note**: Before running any code set the right **Working Directory** and Change the file path accordingly.

#ANALYSIS

####Install required packages

"tidyverse" for data import and wrangling
"libridate" for date functions
"ggplot" for visualization


```{r Load libraries}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
getwd() #displays your working directory
setwd("D:/Capstone_Project/R Analysis/Extracted") #sets your working directory to simplify calls to data.

```


## STEP 1: COLLECT DATA

### Upload Divvy datasets (csv files) here
```{r Import Data}
q2_2019 <- read.csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read.csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read.csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read.csv("Divvy_Trips_2020_Q1.csv")

```


## STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE

Compare column names each of the files
While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file

```{r Check Columns}
colnames(q3_2019)
colnames(q4_2019)
colnames(q2_2019)
colnames(q1_2020)

```

Rename columns  to make them consisent with q1_2020

```{r Rename}
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q2_2019 <- rename(q2_2019
                   ,ride_id = "X01...Rental.Details.Rental.ID"
                   ,rideable_type = "X01...Rental.Details.Bike.ID"
                   ,started_at = "X01...Rental.Details.Local.Start.Time" 
                   ,ended_at = "X01...Rental.Details.Local.End.Time"  
                   ,start_station_name = "X03...Rental.Start.Station.Name" 
                   ,start_station_id = "X03...Rental.Start.Station.ID"
                   ,end_station_name = "X02...Rental.End.Station.Name" 
                   ,end_station_id = "X02...Rental.End.Station.ID"
                   ,member_casual = "User.Type"))

```

Inspect the dataframes and look for incongruencies

```{r Inspect}
str(q1_2020)
str(q4_2019)
str(q3_2019)
str(q2_2019)

```
Convert ride_id and rideable_type to character so that they can stack correctly

```{r datatypes}

q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 

```

Stack individual quarter's data frames into one big data frame

```{r Stack}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)

```

Remove lat, long, birthyear, and gender fields as this data was dropped beginning in 2020

```{r Remove}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "X01...Rental.Details.Duration.In.Seconds.Uncapped", "X05...Member.Details.Member.Birthday.Year", "Member.Gender", "tripduration"))

```


## STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS

Inspect the new table that has been created

```{r Inspect New}

colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(qs_raw)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics

```

There are a few problems we will need to fix:
(1) In the "member_casual" column, there are two names for members ("member" and "Subscriber") and two names       for casual riders ("Customer" and "casual"). We will need to consolidate that from four to two labels.
(2) The data can only be aggregated at the ride-level, which is too granular. We will want to add some             additional columns of data -- such as day, month, year -- that provide additional opportunities to             aggregate the data.
(3) We will want to add a calculated field for length of ride since the 2020Q1 data did not have the               "tripduration" column. We will add "ride_length" to the entire dataframe for consistency.
(4) There are some rides where tripduration shows up as negative, including several hundred rides where Divvy      took bikes out of circulation for Quality Control reasons. We will want to delete these rides.

In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual"
Before 2020, Divvy used different labels for these two types of riders ... we will want to make our dataframe consistent with their current nomenclature

Begin by seeing how many observations fall under each usertype

```{r User Type}
table(all_trips$member_casual)

```

Reassign to the desired values 

```{r Reassign}
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))

```

Check to make sure the proper number of observations were reassigned

```{r Make Sure}
table(all_trips$member_casual)

```

Add columns that list the date, month, day, and year of each ride
This will allow us to aggregate ride data for each month, day, or year

```{r Add Columns}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

Add a "ride_length" calculation to all_trips (in seconds)

```{r Add Calculated Column}

all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

```

Inspect the structure of the columns

```{r Insppect Structure}
str(all_trips)

```

Convert "ride_length" from Factor to numeric so we can run calculations on the data

```{r Convert}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)

```

Remove "bad" data
The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative

We will create a new version of the dataframe (v2) since data is being removed

```{r Create New}

all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

```


## STEP 4: CONDUCT DESCRIPTIVE ANALYSIS

Descriptive analysis on ride_length (all figures in seconds)

```{r Analysis}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride

```

Or we can use the summary function

```{r Summary}
summary(all_trips_v2$ride_length)

```


Compare members and casual users

```{r Members vs Casual users}

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```

See the average ride time by each day for members vs casual users

```{r average ride time by each day for members vs casual users}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```


Notice that the days of the week are out of order. Let's fix that.

```{r Fix}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

Now, let's run the average ride time by each day for members vs casual users

```{r average ride time by each day for members vs casual users 2}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

analyze ridership data by type and weekday

```{r Type and Weekday}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)								# sorts
```

### Let's visualize the number of rides by rider type
```{r rides by rider type}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

### Let's create a visualization for average duration
```{r average duration}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge") + 
  labs(title = "average duration of casual_members and subscribers")
```


## STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS

#### Create a csv file that we will visualize in Excel, Tableau, or other presentation software
```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = 'D:/Capstone_Project/R Analysis/Extracted/avg_ride_length.csv')
```
